### 2020-11-18 popl21/livelit-calculus.tex
moved popl21 submission into a directory for archival

\section{A Simply Typed Livelit Calculus}\label{sec:livelit-calculus}
In order to precisely capture the semantics of livelits
independently
of the specifics of the Hazel programming environment and GUI
framework, we will now specify a simply typed \emph{livelit calculus}.

Fig.~\ref{fig:syntax} specifies the syntax of the livelit calculus.
Programs are written as \emph{unexpanded expressions}, $\hat e$, which are \emph{expanded} to
\emph{external expressions}, $e$, before being \emph{elaborated}
to internal expressions, $d$, for evaluation. All three stages are governed
by the same types, $\tau$. We include numbers, partial functions, products, sums, and recursive
types, all in their standard form \cite{pfpl}, but this specific type structure is not critical.
Any language expressive
enough to encode its own syntax (here as values of a recursive sum type)
would be a suitable basis. We begin with a terse
overview of the external and internal languages,
which are adapted straightforwardly from Hazelnut Live \cite{HazelnutLive} and therefore
serve as background material, in Sec.~\ref{sec:external-and-internal-lang}.
Only unexpanded expressions contain
livelits, so they are the main focus of the remaining sections,
which cover livelit expansion (Sec.~\ref{sec:calculus-expansion}),
livelit update actions (Sec.~\ref{sec:calculus-updates}),
liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}),
and parameterization (Sec.~\ref{sec:calculus-parameterization}).

\subsection{Background: External and Internal Language}\label{sec:external-and-internal-lang}
The external and internal languages are straightforward adaptations of the
external and internal languages of \emph{Hazelnut Live},
a typed lambda calculus that assigns static and dynamic meaning to programs with holes,
notated externally $\hehole{u}$ where $u$ is a \emph{hole name} \cite{HazelnutLive}.
We omit non-empty holes (which allow reasoning about type inconsistencies \cite{Hazelnut}) and type holes
(which operate like the unknown type from gradual type theory \cite{Siek06a,Hazelnut})
because these mechanisms are orthogonal to livelits.

@@

\subsection{Expansion}

    \begin{figure}
    \begin{mathpar}
        \inferrule[EVar]{
            x : \tau \in \Gamma
        }{
            \expands{\Phi}{\Gamma}{x}{x}{\tau}
        }

@@

        \inferrule[EApLivelit]{
            \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi\\\\
            \hasType{ }{\dtxt{model}}{\taut{model}}\\
            \evalsTo{\hap{\dtxt{expand}}{\dtxt{model}}}{\dtxt{encoded}}\\
            \decodeExp{\dtxt{encoded}}{\etxt{pexpansion}}\\\\
            \hasType{ }{\etxt{pexpansion}}{\tarr{\{\tau_i\}_{i < n}}{\taut{expand}}}\\
            \{ \expands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i} \}_{i < n}
        }{
            \expands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{
                \hap{\etxt{pexpansion}}{\{e_i\}_{i < n}}
            }{\taut{expand}}
        }
    \end{mathpar}
    \caption{Expansion}
    \label{fig:expansion}
    \end{figure}
The novelty of the livelit calculus lies entirely in its handling of unexpanded expressions,
$\hat e$, which are given meaning by typed expansion to external expressions,
$e$, according to the judgement $\expands{\Phi}{\Gamma}{\hat e}{e}{\tau}$
defined in Fig.~\ref{fig:expansion}.
Unexpanded expressions mirror external expressions but for the presence of livelits, discussed below.
The rules for standard
forms like variables, functions and function application, shown in Fig.~\ref{fig:expansion},
are straightforward.

\subsubsection{Livelit Contexts}
Livelit definitions are collected in the livelit context, $\Phi$, which
maps livelit names to livelit definitions of the following form:
\[ \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \]
Here, $\taut{expand}$ is the expansion type, $\taut{model}$ is the model type,
and $\dtxt{expand}$ is the expansion function, which generates an expansion given a model.\todo{add view and update}

For simplicity, we assume that the livelit context is provided \emph{a priori} and therefore
that functions like $\dtxt{expand}$ are already closed and fully elaborated.
In practice, the livelit context would be controlled by a definition form in the language
that allows the expansion function to itself use livelits.
This would require a staging mechanism,
because we need to execute expansion functions in prior definitions to be able to
expand subsequent definitions.
There are a number of ways to support the necessary staging in practice, e.g.
via explicit staging primitives \todo{cite ocaml macros paper, Racket stuff}{},
by requiring that these definitions appear in separately compiled packages \cite{TLMs},
or by using live
programming mechanisms such as those available in Hazel
to evaluate ``up to'' each definition before proceeding.
This choice is orthogonal to the mechanisms of interest in this section.

\subsubsection{Hygienic Livelit Expansion}
Unexpanded expressions are unique in that they include livelits:
 \[\haplivelit{u}{a}{\dtxt{model}}{\splices}\]
Here, $\livelitname{a}$ names the livelit
 being applied. Livelits can be understood as filling holes, so $u$ identifies the hole
 that is, conceptually, being filled.
The current state of the livelit is determined by the current model value, $\dtxt{model}$,
together with a collection of splices, $\splices$. Each splice $\psi_i$
is of the form $\hsplice{\hat e_i}{\tau_i}$, where $\hat e_i$ is the spliced expression
itself (unexpanded, so it may contain other livelits) and $\tau_i$ is the type of that splice,
as determined when the livelit definition first requested the splice (discussed in Sec.~\ref{sec:calculus-updates}).

Rule \rulename{EApLivelit} performs livelit expansion. Its premises, in order, operate as follows:
\begin{enumerate}
    \item \textbf{Lookup.} The first premise looks up the livelit name in the livelit context.
    \item \textbf{Model Validation.} The second premise serves to ensure that the model value, $\dtxt{model}$, is of the
    specified model type, $\taut{model}$.
    \item \textbf{Expansion.} The third premise applies the expansion function, $\dtxt{expand}$, to the model value, $\dtxt{model}$,
    producing the encoded expansion, $\dtxt{encoded}$, which, by the definitions given above, is of type $\mathsf{Exp}$.
    \item \textbf{Decoding.} The fourth premise decodes $\dtxt{encoded}$, producing the external expression it encodes, $\etxt{pexpansion}$.
    \item \textbf{Expansion Validation.} We call $\etxt{expansion}$ the \emph{parameterized expansion} because, according to the fifth premise,
    it must be a function that returns a value of the expansion type, $\taut{expand}$, when applied (in curried fashion, though this is not critical)
    to the splices, whose types, $\splat{\tau_i}$, are explicitly recorded in the livelit.
    We must take care to ensure that the parameterized expansion is \emph{context independent},
    i.e. that it cannot depend on the particular bindings available in the call site typing context, $\Gamma$.
    Context independence is one facet of what is colloquially known as \emph{hygiene} \cite{TLMs}.
    In this simple formulation of the system, we maintain context independence simply by
    requiring that the parameterized expansion be
    closed. Consequently, any necessary helper functions used in the expansion must be provided explicitly by the client
    via a splice. In Sec.~\ref{sec:calculus-parameterization}, we will see how this context independence
    restriction can be made
    less onerous by passing definition-site bindings along as partially applied parameters.\todo{remove?}{}
    \item \textbf{Splice Expansion.} Finally, the sixth premise inductively expands each of the spliced expressions in the same context as the livelit
    expression itself.
\end{enumerate}
The conclusion of the rule then applies the parameterized expansion to the expanded splices.
By applying the splices as arguments, we maintain \emph{capture avoidance} -- splices cannot capture variables
bound internally to the expansion. Capture avoidance is the other facet of \emph{hygiene} \cite{TLMs}.

The typed expansion process is governed by the following metatheorem, which establishes that the expansion
is indeed an external expression of the indicated type.\todo{change to proof sketch}{}

@@

When composed with the Typed Elaboration theorem and the Type Safety properties established for the internal
language, we achieve end-to-end type safety for unexpanded expressions (every well-typed unexpanded
expression expands to a well-typed external expression, which in turn expands to a well-typed internal
expression, which evaluates in a type safe manner.)

\subsubsection{Agda Mechanization}
We have mechanized the typed expansion mechanism and proven the Typed Expansion theorem
using the Agda proof assistant.\todo{say more}{}

\subsection{Live Feedback via Closure Collection}
In order to support live feedback, a livelit needs to be able to ask the system
to evaluate expressions under one of the closures associated with the livelit.
This mechanism was introduced by example in Sec.~\ref{sec:image-transformation}.
In this section, we will formalize the process of efficiently collecting closures
for the livelits that appear in a program.

@@

Consequently, the second step of closure selection is to fill any livelit holes that appear in the proto-environments for other livelit holes.
We do so by filling them using the parameterized expansions gathered in $\Omega$ and then resuming
evaluation where appropriate.
Formally, this involves the hole filling operation $\instantiate{d_1}{u}{d_2}$ for Hazelnut Live
(which derives from the metavariable instantation operation of contextual modal type theory \cite{HazelnutLive,Nanevski2008}).
This operation
fills every closure for hole $u$ in $d_2$ with $d_1$.
Unlike substitution, hole filling is
not capture-avoiding. Instead, the environment on each of these closures is applied to $\dtxt{pexpansion}$
as a substitution, i.e. the delayed substitutions captured in the environment are realized.
In this case, however, the parameterized expansion is guaranteed to be closed due to
the context independence discipline we maintain in Rule \rulename{EApLivelit},
so hole filling acts simply as a syntactic replacement.

@@

\subsection{Parameterization}

Add livelit expressions, livelit abbreviations,
partial application support. Talk about how to work around the strict context
independence discipline we have imposed.

\subsection{Edit Action Semantics}
So far, we have considered only the semantics of an individual editor state.
However, livelits are useful because they evolve as the user interacts
with them. In particular, the livelit's model, $\dtxt{model}$ can evolve
due to livelit-specific user interactions. Furthermore, a livelit can also
create and delete splices when a user interaction demands it (unlike functions, which have a fixed
number of arguments).

>>>>

\section{A Simply Typed Livelit Calculus}\label{sec:livelit-calculus}

In order to precisely capture the semantics of livelits
independently
of the specifics of the Hazel programming environment and GUI
framework, we will now specify a simply typed \emph{livelit calculus}.

Fig.~\ref{fig:syntax} specifies the syntax of the livelit calculus.
Programs are written as \emph{unexpanded expressions}, $\hat e$, which are \emph{expanded} to
\emph{external expressions}, $e$, before being \emph{elaborated}
to internal expressions, $d$, for evaluation. All three stages are governed
by the same types, $\tau$. We include numbers, partial functions, products, sums, and recursive
types, all in their standard form \cite{pfpl}, but this specific type structure is not critical.
Any language expressive
enough to encode its own syntax (here as values of a recursive sum type)
would be a suitable basis. We begin with a terse
overview of the external and internal languages,
which are adapted straightforwardly from Hazelnut Live \cite{HazelnutLive} and therefore
serve as background material, in Sec.~\ref{sec:external-and-internal-lang}.
Only unexpanded expressions contain
livelits, so they are the main focus of the remaining sections,
which cover livelit expansion (Sec.~\ref{sec:calculus-expansion}) and 
liveness via closure collection (Sec.~\ref{sec:calculus-closure-collection}). 
% and parameterization (Sec.~\ref{sec:calculus-parameterization}).

\subsection{Background: External and Internal Language}\label{sec:external-and-internal-lang}
The external and internal languages are straightforward adaptations of the
external and internal languages of \emph{Hazelnut Live},
a typed lambda calculus that assigns static and dynamic meaning to programs with holes,
notated externally $\hehole{u}$ where $u$ is a \emph{hole name} \cite{HazelnutLive}.
We omit non-empty holes (which internalize type inconsistencies \cite{Hazelnut}) and type holes
(which operate like the unknown type from gradual type theory \cite{Siek06a,Hazelnut})
because these mechanisms are orthogonal to livelits.

@@

\subsection{Expansion}\label{sec:calculus-expansion}

The novelty of the livelit calculus lies entirely in its handling of unexpanded expressions,
$\hat e$, which are given meaning by typed expansion to external expressions,
$e$, according to the judgement $\expands{\Phi}{\Gamma}{\hat e}{e}{\tau}$
defined in Fig.~\ref{fig:expansion}.
Unexpanded expressions mirror external expressions but for the presence of livelits, discussed below.
The rules for standard
forms like variables, functions and function application, shown in Fig.~\ref{fig:expansion},
are straightforward.

\subsubsection{Livelit Contexts}

Livelit definitions are collected in the livelit context, $\Phi$, which
maps livelit names to livelit definitions of the following form:
\[ \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \]
Here, $\taut{expand}$ is the expansion type, $\taut{model}$ is the model type,
and $\dtxt{expand}$ is the expansion function, which generates an expansion given a model.
We omit logic related to view computations and actions, which are tied to a particular 
UI framework.

For simplicity, we assume that the livelit context is provided \emph{a priori} and therefore
that functions like $\dtxt{expand}$ are already closed and fully elaborated.
In practice, the livelit context would be controlled by a definition form in the language
that allows the expansion function to itself use livelits.
This would require a staging mechanism,
because we need to execute expansion functions in prior definitions to be able to
expand subsequent definitions.
There are a number of ways to support the necessary staging in practice, e.g.
via explicit staging primitives \cite{DBLP:conf/icfp/Flatt02},
by requiring that these definitions appear in separately compiled packages \cite{TLMs},
or by using live
programming mechanisms such as those available in Hazel
to evaluate ``up to'' each definition before proceeding \cite{HazelnutLive}.
This choice is orthogonal to the mechanisms of interest in this section.

\subsubsection{Hygienic Livelit Expansion}
\begin{figure}
    \begin{mathpar}
        \inferrule[EVar]{
            x : \tau \in \Gamma
        }{
            \expands{\Phi}{\Gamma}{x}{x}{\tau}
        }

@@

        \inferrule[EApLivelit]{
            \livelitCtxEntry{a}{\taut{expand}}{\taut{model}}{\dtxt{expand}} \in \Phi\\\\
            \hasType{ }{\dtxt{model}}{\taut{model}}\\
            \evalsTo{\hap{\dtxt{expand}}{\dtxt{model}}}{\dtxt{encoded}}\\
            \decodeExp{\dtxt{encoded}}{\etxt{pexpansion}}\\\\
            \hasType{ }{\etxt{pexpansion}}{\tarr{\{\tau_i\}_{i < n}}{\taut{expand}}}\\
            \{ \expands{\Phi}{\Gamma}{\hat e_i}{e_i}{\tau_i} \}_{i < n}
        }{
            \expands{\Phi}{\Gamma}{\haplivelit{u}{a}{\dtxt{model}}{\splat{\hsplice{\hat e_i}{\tau_i}}}}{
                \hap{\etxt{pexpansion}}{\{e_i\}_{i < n}}
            }{\taut{expand}}
        }
    \end{mathpar}
    \caption{Expansion}
    \label{fig:expansion}
    \end{figure}
Unexpanded expressions are unique in that they include livelits:
 \[\haplivelit{u}{a}{\dtxt{model}}{\splices}\]
Here, $\livelitname{a}$ names the livelit
 being applied. Livelits can be understood as filling holes, so $u$ identifies the hole
 that is, conceptually, being filled.
The current state of the livelit is determined by the current model value, $\dtxt{model}$,
together with a collection of splices, $\splices$. Each splice $\psi_i$
is of the form $\hsplice{\hat e_i}{\tau_i}$, where $\hat e_i$ is the spliced expression
itself (unexpanded, so it may contain other livelits) and $\tau_i$ is the type of that splice,
as determined when the livelit definition first requested the splice (discussed in Sec.~\ref{sec:livelit-definitions}).

Rule \rulename{EApLivelit} performs livelit expansion. Its premises, in order, operate as follows:
\begin{enumerate}
    \item \textbf{Lookup.} The first premise looks up the livelit name in the livelit context.
    \item \textbf{Model Validation.} The second premise serves to ensure that the model value, $\dtxt{model}$, is of the
    specified model type, $\taut{model}$.
    \item \textbf{Expansion.} The third premise applies the expansion function, $\dtxt{expand}$, to the model value, $\dtxt{model}$,
    producing the encoded expansion, $\dtxt{encoded}$, which, by the definitions given above, is of type $\mathsf{Exp}$.
    \item \textbf{Decoding.} The fourth premise decodes $\dtxt{encoded}$, producing the external expression it encodes, $\etxt{pexpansion}$.
    \item \textbf{Expansion Validation.} We call $\etxt{expansion}$ the \emph{parameterized expansion} because, according to the fifth premise,
    it must be a function that returns a value of the expansion type, $\taut{expand}$, when applied (in curried fashion, though this is not critical)
    to the splices, whose types, $\splat{\tau_i}$, are explicitly recorded in the livelit.
    We must take care to ensure that the parameterized expansion is \emph{context independent},
    i.e. that it cannot depend on the particular bindings available in the call site typing context, $\Gamma$.
    Context independence is one facet of what is colloquially known as \emph{hygiene} \cite{TLMs}.
    In this simple formulation of the system, we maintain context independence simply by
    requiring that the parameterized expansion be
    closed. Consequently, any necessary helper functions used in the expansion must be provided explicitly by the client
    via a splice. In Sec.~\ref{sec:livelit-definitions}, we discussed how the use of explicit capture sets, which can be understood as 
    immediate partial application, can lower the cognitive burden of this strict context independence discipline.
    Parameters formally operate in the same way as splices, and partial application is a well-understood technique, so we omit these details from the calculus.
    \item \textbf{Splice Expansion.} Finally, the sixth premise inductively expands each of the spliced expressions in the same context as the livelit
    expression itself.
\end{enumerate}
The conclusion of the rule then applies the parameterized expansion to the expanded splices.
By applying the splices as arguments, we maintain \emph{capture avoidance} -- splices cannot capture variables
bound internally to the expansion. Capture avoidance is the other facet of \emph{hygiene} \cite{TLMs}.

The typed expansion process is governed by the following metatheorem, which establishes that the expansion
is indeed an external expression of the indicated type.

@@

When composed with the Typed Elaboration theorem and the Type Safety properties established for the internal
language, we achieve end-to-end type safety for unexpanded expressions (every well-typed unexpanded
expression expands to a well-typed external expression, which in turn expands to a well-typed internal
expression, which in turn evaluates in a type safe manner.)

\subsubsection{Agda Mechanization}
We have mechanized the typed expansion mechanism and proven the Typed Expansion theorem
using the Agda proof assistant, using the same techniques as were used in the Agda mechanization 
of Hazelnut Live \cite{HazelnutLive}. This mechanization is available in the anonymous 
supplemental material.

\subsection{Live Feedback via Closure Collection}\label{sec:calculus-closure-collection}
In order to support live feedback, a livelit needs to be able to ask the system
to evaluate expressions under one of the closures associated with the livelit.
This mechanism was introduced by example in Sec.~\ref{sec:image-transformation}.
In this section, we will formalize the process of efficiently collecting closures
for the livelits that appear in a program.

@@

Consequently, the second step of closure selection is to fill any livelit holes that appear in the proto-environments for other livelit holes.
We do so by filling them using the parameterized expansions gathered in $\Omega$ and then resuming
evaluation where appropriate.
Formally, this involves the hole filling operation $\instantiate{d_1}{u}{d_2}$ for Hazelnut Live
(which derives from the metavariable instantation operation of contextual modal type theory \cite{HazelnutLive,Nanevski2008}).
This operation
fills every closure for hole $u$ in $d_2$ with $d_1$.
Unlike substitution, hole filling is
not capture-avoiding. Instead, the environment on each of these closures is applied to $d_1$
as a substitution, i.e. the delayed substitutions captured in the environment are realized.
In this case, however, the parameterized expansion is guaranteed to be closed due to
the context independence discipline we maintain in Rule \rulename{EApLivelit},
so hole filling acts simply as a syntactic replacement.

@@

% \subsection{Parameterization}

% Add livelit expressions, livelit abbreviations,
% partial application support. Talk about how to work around the strict context
% independence discipline we have imposed.

% \subsection{Edit Action Semantics}
% So far, we have considered only the semantics of an individual editor state.
% However, livelits are useful because they evolve as the user interacts
% with them. In particular, the livelit's model, $\dtxt{model}$ can evolve
% due to livelit-specific user interactions. Furthermore, a livelit can also
% create and delete splices when a user interaction demands it (unlike functions, which have a fixed
% number of arguments).
